Geometrization of characters of tori over non-Archimedean local fields

### 2013-08-13 shanghai.tex
Added a paragraph at the end of the snake section to explain why neither finite nor bounded geochars can be used to geometrize characters of G(Fq).

@@

\section{The category of geometrized characters}\label{sec:category}

@@

\section{Pull-back of geometrized characters}\label{subsection:pullback}

@@

\begin{proof}
The only non-trivial part is essential surjectivity. Set $G \ceq G_1\times G_1$. Let $\cs{L}$ be a geochar on $G$. Define $f : G\to G\times G$ by $f(g_1,g_2) = (g_1,e_2,e_1,g_2)$. Observe that $m\circ f = \id_G$. 
Then the isomorphism $\mu : m^* \gcs{L} \to \gcs{L}\boxtimes \gcs{L}$ gives an isomorphism
\[
f^*\mu : f^* m^* \gcs{L} \to f^*(\gcs{L}\boxtimes \gcs{L}) = f^*p_1^*\gcs{L}\otimes f^* p_2^*\gcs{L}.
\]
Observe that $p_1\circ f = r_1 \circ q_1$ and $f_2 = r_2\circ f$ where $p_1 , p_2 : G\times G \to G$ are the usual things and $r_1 : G_1\times G_2 \to G_1$ and $r_2 : G_1\times G_2 \to G_2$ are the obvious things and $q_1 : G_1\to G_1\times G_2$ and  $q_2 : G_2\to G_1\times G_2$ are the usual things. Then $f^*p_1^*\gcs{L} = r_1^* q_1^* \gcs{L}$ and $f^*p_2^*\gcs{L} = r_2^* q_2^* \gcs{L}$. Now $\cs{L}_1 \ceq r_1^* \cs{L}$ is a geochar on $G_1$ and $\cs{L}_2 \ceq r_2^* \cs{L}$ is a geochar on $G_2$, by Proposition~\ref{prop:pullback}. Then
$ f^*p_1^*\gcs{L}\otimes f^* p_2^*\gcs{L} = r_1^*\gcs{L}_1\otimes r_2^* \gcs{L}_2 = \gcs{L}_1\boxtimes \gcs{L}_2$. In this way, $\mu$ determines an isomorphism
\[
\gamma \ceq f^*\mu : \gcs{L} \to  \gcs{L}_1\boxtimes \gcs{L}_2.
\]
So $\boxtimes$ is essentially surjective.
\end{proof}

\section{Base change over the residue field}\label{sec:basechange}

@@

We will see that the full subcategory $\GC_0(G)$ of bounded geometrized characters is a proper subcategory of $\GC(G)$, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to G(\Fq)^*$ to the subgroup $\GCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is bounded in $\EEx$, so if $G(\Fq)$ is not finite, then $\GC_0(G) \hookrightarrow\GC(G)$ is not essentially surjective.

@@

We will see that the full subcategory $\GC_f(G) \subset \GC_0(G)$ of finite geometrized characters is a proper subcategory, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to G(\Fq)^*$ to the subgroup $\GCfiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is finite in $\EEx$.

@@

\section{Geometrized characters of commutative \'etale group schemes}

@@

\section{Main result regarding geometrized characters}

@@

\Clifton{Return to finite and bounded character sheaves, explaining why our definition is the right one. Also explain why we don't think of local systems as representations of a fundamental group.}

>>>>

Acknowledgements: Masoud Kamgarpour, Pramod Achar, Hadi Salmasian.

@@

\section{Geometrized characters}\label{sec:category}

@@

\section{Pull-back and products}\label{subsection:pullback}

@@

\begin{proof}
The only non-trivial part is essential surjectivity. Set $G \ceq G_1\times G_1$. Let $\cs{L}$ be a geochar on $G$. Define $f : G\to G\times G$ by $f(g_1,g_2) = (g_1,e_2,e_1,g_2)$. Observe that $m\circ f = \id_G$. 
Then the isomorphism $\mu : m^* \gcs{L} \to \gcs{L}\boxtimes \gcs{L}$ gives an isomorphism
\[
f^*\mu : f^* m^* \gcs{L} \to f^*(\gcs{L}\boxtimes \gcs{L}) = f^*p_1^*\gcs{L}\otimes f^* p_2^*\gcs{L}.
\]
Observe that $p_1\circ f = r_1 \circ q_1$ and $f_2 = r_2\circ f$ where $p_1 , p_2 : G\times G \to G$ are the usual things and $r_1 : G_1\times G_2 \to G_1$ and $r_2 : G_1\times G_2 \to G_2$ are the obvious things and $q_1 : G_1\to G_1\times G_2$ and  $q_2 : G_2\to G_1\times G_2$ are the usual things. Then $f^*p_1^*\gcs{L} = r_1^* q_1^* \gcs{L}$ and $f^*p_2^*\gcs{L} = r_2^* q_2^* \gcs{L}$. Now $\cs{L}_1 \ceq r_1^* \cs{L}$ is a geochar on $G_1$ and $\cs{L}_2 \ceq r_2^* \cs{L}$ is a geochar on $G_2$, by Proposition~\ref{prop:pullback}. Then
$ f^*p_1^*\gcs{L}\otimes f^* p_2^*\gcs{L} = r_1^*\gcs{L}_1\otimes r_2^* \gcs{L}_2 = \gcs{L}_1\boxtimes \gcs{L}_2$. In this way, $\mu$ determines an isomorphism
\[
f^*\mu : \gcs{L} \to  \gcs{L}_1\boxtimes \gcs{L}_2.
\]
So $\boxtimes$ is essentially surjective.
\end{proof}

\section{Base change}\label{sec:basechange}

@@

The full subcategory $\GCb(G)$ of bounded geometrized characters is an essentially proper subcategory of $\GC(G)$, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to G(\Fq)^*$ to the subgroup $\GCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is bounded in $\EEx$; see Section~\ref{sec:snake}.

@@

The full subcategory $\GCf(G) \subset \GCb(G)$ of finite geometrized characters is a proper subcategory, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to G(\Fq)^*$ to the subgroup $\GCfiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is finite in $\EEx$; see Section~\ref{sec:snake}.

@@

\section{Commutative \'etale group schemes}

@@

\section{Main result regarding geometrized characters}
\label{sec:snake}

@@

With Theorem~\ref{thm:snake} on hand, we can see why neither bounded geochars nor finite geochars are up to the task of geometrizing characters of $G(\Fq)$ in the generality we need.  To see this, consider the case when $G$ is the discrete etale group scheme $\ZZ$. Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ and let $\cs{L}$ be a geometrized character in the isomorphism class corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. Then $\trFrob{\cs{L}}(1) = \ell$. Suppose $\cs{L}$ was a bounded character sheaf. Then the restriction of $\cs{L}$ to the connected component $1\in G$ would be a rank-1 local system on $1$, thus an $\ell$-adic chacter $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with $\rho_1(\Frob{\Fq}) = \ell$. But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, while $\rho_1(\pi_1(1,{\bar 1}))$ is not compact (it contains $\ell^\ZZ$). Since the Weil sheaf $\gcs{L}$ does not descend to a local system on $G$, the geometric character $\cs{L}$ is not a bounded geometric character. We can play the same game with the character $\chi : G(\Fq) \to \EEx$ defined by $\chi(1) = 1+\ell$. In this case, the corresponding geometric character $\cs{L}$ is bounded, but not finite. This simple example illustrates why the categories $\GCb(G)$ and $\GCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.
